---
title: "Q2"
author: "Anouk Dimbeanu"
date: "2025-03-30"
output: html_document
---

```{r}
library(dplyr)
library(tidyr)
library(caret)
library(MASS) 
url <- "https://raw.githubusercontent.com/emmaamitchyy/BIOL432_Gr3_Final/refs/heads/main/Data_Input/RawData.csv"
df <- read.csv(url)

```

```{r}
df$Age <- as.numeric(df$Age)
df$CT <- as.numeric(df$CT)
df$Age[is.na(df$Age)] <- mean(df$Age, na.rm = TRUE) 
head(df)

```

## Change the characters to numeric values

```{r}
dat <- df %>%
  mutate(across(c(where(is.character), -c(Sample.Name, Batch.Number, Class.name, Sex, CT, Age)), as.numeric)) %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0))) 
```


## Rearrange dataset 

```{r}
dat <- as_tibble(dat) 
dat <- dat[, c("Sample.Name", "Sex", "Age", "CT", setdiff(names(dat), c("Sample.Name", "Sex", "Age", "CT")))]
head(dat)
```

## Feature selection for only COVID-19


```{r}
respDat <- dat %>% 
  dplyr::select(1:6)
Features <- dat %>% 
  dplyr::select(-c(1:6))
```

## Scaling

```{R}
Scaled<-Features %>%mutate_all(scale)
```


## Find missing data

```{r}
Scaled %>%
  select_if(function(x) any(is.na(x))) %>%
  names()
```
 
 We only have to change Trp

```{r}
ScalComp <- Scaled %>% 
  mutate(Trp = ifelse(is.na(Trp), 0, Trp))
```


## Dimension reduction

```{r}
dim(ScalComp)
```

## PCA

```{r}
PCA <- princomp(ScalComp)
summary(PCA)
```

## Feature selection 

```{r}
FeatureSel <- ScalComp %>%
  mutate(Class.name = dat$Class.name) %>%
  pivot_longer(cols = -Class.name,
               names_to = "metabolite",
               values_to = "CONC")
str(FeatureSel) 
```


```{r}
heatmapdata <- ScalComp %>%
  mutate(Class.name = dat$Class.name) %>%
  filter(Class.name == "COVID19") %>%
  as.data.frame()
heatmapdata$CT <- as.numeric(dat$CT[match(rownames(heatmapdata), rownames(dat))])
heatmapdata$CT[is.na(heatmapdata$CT)] <- 0
head(heatmapdata)
```

```{r}
heatmapdata <- heatmapdata[, !names(heatmapdata) %in% "Class.name"]
head(heatmapdata)
```

```{r}
heatmapdata$CT <- make.unique(as.character(heatmapdata$CT))
rownames(heatmapdata) <- heatmapdata$CT
heatmapdata <- heatmapdata[, -which(names(heatmapdata) == "CT")]
heatmapdata <-heatmapdata[,-1]
heatmapdata<-as.matrix(heatmapdata)
heatmapdata
```


```{r}
library(pheatmap)
heatmapdata <- heatmapdata[rev(rownames(heatmapdata)), ]
pheatmap(heatmapdata, cluster_rows = F, cluster_cols = F, 
         color = colorRampPalette(terrain.colors(8))(100),
         fontsize_row = 4,
         fontsize_col = 5,
         xlab = "Metabolites",
         ylab = "CT values",       
         main = "COVID-19 metabolites correlation with CT",
         legend_labels = c("Low", "Medium", "High"))
```
Figure 1. Heatmap of all the metabolites and their correlation by cycle threshold of COVID-19

## Find the top 10 scoring metabolites and plot them in a heatmap

```{r}
topscores <- apply(heatmapdata, 2, max)  
topmetabolites <- names(sort(topscores))[1:10]  

top_10_data <- heatmapdata[, topmetabolites]
pheatmap(top_10_data, cluster_rows = F, cluster_cols = F, 
         color = colorRampPalette(terrain.colors(8))(100),
         fontsize_row = 4,
         fontsize_col = 9,
         xlab = "Metabolites",
         ylab = "CT values",       
         main = "COVID-19 metabolites correlation with CT",
         legend_labels = c("Low", "Medium", "High"))
```
Figure 2. Top 10 scoring metabolites and their correlation by cycle threshold for COVID-19

Pheatmap code taken from rdocumentation:
https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12/topics/pheatmap


