---
title: 'BIOL 432: Final Project'
author: 
- "Emma Mitchell - 20296602"
- "Grace Wolfe - 20302888"
- "Autumn Hodgins - 20270064"
- "Noah Gandl Black - 20301317"
- "Anouk Dimbeanu - 20265582"
- "Cameron Debellefeuille - 20324416"

- 
date: "2025-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BIOL 432 - Group 3 Final Project 

# Analysis of The Nasal Metabolome: 

Git: https://github.com/emmaamitchyy/BIOL432_Gr3_Final 

 # Introduction: 

The COVID-19 pandemic, caused by the SARS-CoV-2 virus, has had profound global health and economic impacts. While extensive research has been conducted on the virus’s transmission, pathology and immune response, we wanted to look at the metabolic alterations associated with infection. Metabolomics, the study of metabolites within biological systems, provides valuable insights into disease mechanisms, potential biomarkers, and therapeutic targets. By examining the nasopharyngeal metabolome, and therapeutic targets we can assess how metabolic profiles vary among infected individuals and whether specific metabolites correlate with disease severity and patient characteristics such as sex. 
In this project, we analyze a dataset containing metabolomic data from patients infected with COVID-19, Influenza A (INFA), and respiratory syncytial virus (RSV). The dataset includes information on metabolite concentrations, cycle threshold (CT) values (which serve as a proxy for viral load and disease severity), sex, and age. Our goal is to investigate how metabolic profiles differ among infected and non-infected individuals, as well as how these profiles correlate with disease severity and sex differences.
Our goal in this project is to analyze this dataset in order to answer the following questions: 

1. Is there a difference in the metabolite profiles between COVID-19 patients and non-COVID-19 patients (healthy patients and other illnesses)?
To address this question, we will use decision trees and random forest models to classify samples based on their metabolite profiles. Feature importance analysis will help identify which metabolites contribute most to distinguishing COVID-19 patients from those with other infections or no infection.

2. How do Metabolites correlate with COVID severity? 

We will generate heatmap to visualize the relationships between metabolite levels and CT values.

3. Is there a difference in metabolite profiles between people of the opposite sex? 
Principal component analysis (PCA) will be used to explore sex-based differences in metabolic profiles. If significant clustering by sex is observed in the PCA plot, this may suggest inherent metabolic differences between male and female patients in response to viral infections.
## Methods: 


Loading in Packages: 
```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(caret)
library(MASS)
library(randomForest)
library(ggplot2)
library(gbm)
library(pROC)
library(gridExtra)
```

Loading in the data: 
```{r}
df <-  read.csv("https://raw.githubusercontent.com/emmaamitchyy/BIOL432_Gr3_Final/refs/heads/main/Data_Input/RawData.csv")
```


#Q1:
Question: Is there a difference in the metabolite profiles between COVID-19 patients and non-COVID-19 patients (healthy patients and other illnesses)?

Hypothesis: COVID-19 patients will have distinct metabolite profiles compared to non-COVID-19 patients, with certain metabolites being elevated due to immune response and inflammation.

Coding for Q1: 
```{r}
#Data Cleaning for Q1: 
df$Age <- as.numeric(df$Age)
df$CT <- as.numeric(df$CT)
df$Age[is.na(df$Age)] <- mean(df$Age, na.rm = TRUE) 
head(df)
```

```{r}
#Convert special entries to NA
df[df == "< 0"] <- NA
df[df == "No Peak"] <- NA
df <- df %>%
  dplyr::select(-Trp)
```

```{r}
# Create COVID vs Non-COVID label
df$Label <- ifelse(grepl("COVID19", df$Class.name, ignore.case = TRUE), "COVID", "Non-COVID")
df$Label <- as.factor(df$Label)

```

```{r}
#Drop metadata columns
drop_cols <- c("Sample.Name", "Batch.Number", "Sex", "Age", "CT", "Class.name")
df <- df[, !(names(df) %in% drop_cols)]
```

```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
#Convert all feature columns to numeric
df[, names(df) != "Label"] <- lapply(df[, names(df) != "Label"], function(x) as.numeric(as.character(x)))
# NAs introduced
```

Dropping data: 
```{r}
#Drop columns with >20% missing values
threshold <- 0.2 * nrow(df)
df <- df[, colSums(is.na(df)) < threshold] 
#Drop rows with missing Label (if any)
df <- df[!is.na(df$Label), ]
```

```{r}
#Impute missing values using column median
for (col in names(df)) {
  if (col != "Label" && is.numeric(df[[col]])) {
    df[[col]][is.na(df[[col]])] <- median(df[[col]], na.rm = TRUE)
  }
}
```

```{r}
#Split data
set.seed(123)

# Create an 80/20 train/test split
train_index <- createDataPartition(df$Label, p = 0.8, list = FALSE)

# Subset the data into training and testing datasets
train_data <- df[train_index, ]   # 80% of the data
test_data <- df[-train_index, ]   # 20% of the data
```

```{r}
#Training Boosting Model
set.seed(123)

covidBoost <- gbm(Label ~ ., data = train_data,
               distribution = "gaussian",
               n.trees = 1000,
               interaction.depth=3,
               shrinkage = 0.05,
               n.minobsinnode = 5,
               cv.folds = 15)

```

```{r}
# Plotting relative influence
covidBoost_inf <- summary(covidBoost)
colnames(covidBoost_inf) <- c("var", "rel.inf")

plot1 <- ggplot(covidBoost_inf %>% top_n(10, rel.inf), 
       aes(x = rel.inf, y = reorder(var, rel.inf))) +
  geom_col(fill = "steelblue") +
  labs(title = "Influential Metabolites",
       x = "Relative Influence",
       y = "Metabolite") +
  theme_classic()
```
Figure 01: 

```{r}
# Confusion Matrix
set.seed(123)
CatDat <- data.frame(Obs = as.factor(test_data$Label),
                     Pred = round(predict(covidBoost, test_data, type = "response"),0))
table(CatDat)
```
### The confusion matrix.... 
```{r}
#Misclassification rate
MisClass <- CatDat %>% 
  filter(as.numeric(Obs) != Pred)

misclassification_rate <- nrow(MisClass) / nrow(CatDat)
print(misclassification_rate)
```

###The misclassification rate is.. this means that... 


```{r}
#Auc:
roc_curve <- roc(as.numeric(CatDat$Obs), as.numeric(CatDat$Pred))
auc(roc_curve)
```

#The area under the curve is the.... this means that ...
```{r}
#Confusion Matrix
set.seed(123)
CatDat1 <- data.frame(Obs = as.factor(train_data$Label),
                     Pred = round(predict(covidBoost, train_data, type = "response"),0))
table(CatDat1)
```
##The confusion matrix ....
```{r}
#AUC
roc_curve1 <- roc(as.numeric(CatDat1$Obs), as.numeric(CatDat1$Pred))
auc(roc_curve1)
```
#The area under...
Creating plot: 
```{r}
#Graph LYSOC18.0 Concentrations
plot2 <- ggplot(df, aes(x = Label, y = LYSOC18.0, fill = Label)) +
  geom_boxplot() +
  labs(y = "LYSOC18.0 Concentration", x = "Diagnosis") +
  theme_classic()
#Arrange Plots
grid.arrange(plot1, plot2, ncol = 2)
```
Figure 02: 

###Intrepretation/discussion



#Q2: 

Question: How do Metabolites correlate with COVID severity?

Hypothesis: Certain metabolites will correlate with COVID-19 severity, with severe cases exhibiting increased levels of metabolites associated with inflammation and immune dysregulation.



#Q3:

Question: Is there a difference in metabolite profiles betwen people of the opposite sex? 

Hypothesis: Male and female patients will have distinct metabolic profiles, with potential differences in metabolites linked to immune response and inflammation.

To test this hypothesis, we must run a PCA of Metabolite Profiles by Sex. 
```{r}
# Load dataset
dat <- "/Users/emmamitchell/Downloads/RawData.csv"
df <- read.csv(dat, stringsAsFactors = FALSE)
```

```{r}

```






