---
title: "Question 3 Data Cleanup"
author: "Autumn Hodgins"
date: "2025-03-26"
output: html_document
---

Inputting the data and checking the layout with head.
```{r}
library(tidyverse)
dat <- read.csv("Data_Input/data_cleaned.csv")
head(dat)
```

Changing the columns Sample.Name, Batch.Number, Class.name, and Sex to factors.
```{r}
dat <- dat %>%
  mutate(across(c(Sample.Name, Batch.Number, Class.name, Sex), as.factor))
head(dat)
```

Creating a new dataframe called no_NA_dat which only contains rows that do not have an NA value for sex or for Age.
```{r}
no_NA_dat <- dat %>%
  filter(!is.na(Sex)) %>% 
  filter(!is.na(Age))
```


Creating a new dataframe called features that consists of only the columns with metabolites and removes any of the columns which have 0 variance. This data frame is then z-scaled.
```{r}
features <- no_NA_dat %>%
  dplyr::select(4:144) %>% 
  select_if(~ var(., na.rm = TRUE) != 0)

features_scaled <- as.data.frame(scale(features))
head(features_scaled)
```

Setting Up to run a PCA analysis
```{r}
pca_full <- princomp(features_scaled, cor = TRUE)
summary(pca_full)
```
The first principal component axis accounts for 32.2% of the variance, and the second axis accounts for 19.5%, explaining most of the variance of the data with only two axes.

Naming columns for the figure
```{r}
colnames(pca_full$scores) <- paste0("PC", 1:ncol(pca_full$scores))
```


Create PCA scores dataframe
```{r}
pca_scores <- as.data.frame(pca_full$scores)
```


Add the "Sex" and "Age" variables to the PCA scores dataframe
```{r}
pca_scores$Sex <- no_NA_dat$Sex
pca_scores$Age <- no_NA_dat$Age
```


Plot PC1 vs PC2, colour coded by Sex:
```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, colour = Sex)) +
  geom_point() +
  theme_minimal() +
  scale_colour_manual(values = c("Male" = "blue", "Female" = "salmon"))
```
Plot PC1 vs PC2, colour coded by Age:
```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, colour = Age)) +
  geom_point() +
  theme_minimal() +
  scale_colour_gradient(low = "yellow", high = "red")
```



